\textsf{Are large language models superhuman chemists?}

### 2024-10-18 src/tex/appendix.tex
cleanup
try to make ready for logprobs figure

@@

\begin{table*}[!t]
    \centering
        \caption{\textbf{Examples for each of the topics considered in the evaluation of the \chembench corpus.} The table shows the percentage of questions in the corpus that belong to each topic, as well as example questions.}
    \scriptsize
    \begin{tabularx}{\textwidth}{p{2.7cm}p{4em}X}

@@

    \end{tabularx}

@@

\end{table*}

@@

\begin{table*}[!t]
    \centering
        \caption{\textbf{Examples for each of the required skills considered in the \chembench corpus.} The table shows the number of questions for each skill and an example question.}
    \scriptsize
    \begin{tabularx}{\textwidth}{p{2.7cm}p{4em}X}
        \toprule
        Required Skills & Number of Questions & Example Question \\
        \midrule
        Knowledge & 770 & Which of the following salts is an acidic salt? \\
        & & A. \ce{NH4Cl} \\
        & & B. \ce{Na2CO3} \\
        & & C. \ce{NaH2PO4} \\
        & & D. \ce{Zn(OH)Cl} \\
        \midrule
        Reasoning & 825 & How can the optical impact of quantum confinement on the electronic structure of a quantum dot be observed experimentally? \\
        & & A. By measuring atomic force microscopy \\
        & & B. By measuring fourier transform infrared spectroscopy \\
        & & C. By measuring photoluminescence spectrum \\
        & & D. By measuring the absorption spectroscopy \\
        \midrule
        Intuition & 1001 & Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
        & & A. [START\_SMILES]CC1(C)Oc2ccc([N+](=O)[O-])cc2[C@@H](N2CCOCC2)[C\@\@H]1O[END\_SMILES] \\
        & & B. [START\_SMILES]Cc1ccccc1N=C(S)N1CCC(NC(=O)c2ccco2)CC1[END\_SMILES] \\
        \midrule
        Calculation & 91 & Given that the average molar mass of the polymer chains in this sample of poly(lactic acid) (PLA) is \SI{595}{g mol^{-1}} using end-group analysis, where \SI{0.1619}{g} of PLA was dissolved in \SI{25}{cm^3} of benzyl alcohol and titrated with \SI{0.0400}{mol dm^{-3}} \ce{KOH} solution. The volume of \ce{KOH} solution required to reach the endpoint was \SI{6.81}{cm^3}. What is the average number of monomer units in each polymer chain of this sample? \\
        \bottomrule
    \end{tabularx}
    \label{tab:chembench_corpus_cognitive}
\end{table*}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/cb_humanset_v3.pdf}
    \caption{\textbf{Composition of the human subset.} The circular plot shows the distribution of topics and required skills in the human subset of the \chembench corpus. The human subset is a representative subset of the full corpus, with a balanced distribution of topics and skills.}
\end{figure}

\subsection{Model performance}
We also evaluated the model performance on the entire \chembench corpus.
\Cref{fig:barplot_all_correct_all_questions} shows the fraction of questions that were answered correctly by the models.
Note that this ranking differs from the one on the \enquote{tiny} subset

@@

\clearpage
\subsection{Implementation}
An overview of the benchmarking pipeline implemented in \chembench is shown in \cref{fig:process}. More detailed information can be found in the online documentation of the \chembench package at \url{https://lamalab-org.github.io/chem-bench/}.
\begin{figure}
    \centering
    \includegraphics[height=.8\textheight]{figures/process.pdf}
    \caption{\textbf{Overview of the benchmarking pipeline implemented in \chembench.} The process begins with JSON files containing task data, which are used to create \texttt{Task} objects and initialize the \texttt{ChemBenchmark}. \texttt{Prompter} objects are then created to handle different types of prompts for instruction-tuned and completion models.
    The \texttt{Task} Iterator differentiates between \gls{mcq} and other question types. For each task type, appropriate prompts are generated and passed to the \gls{llm}. The responses are then processed using regex-based parsing methods specific to \gls{mcq} or numeric answers (after obtaining the relevant part of the response from the instruction-tuned models).
    The regex-based parsing is elaborate and also able to handle special cases such as scientific notation, or roman numerals.
    If the initial parsing is unsuccessful, the system employs an \gls{llm}-based extraction method as a fallback. The parsed or extracted answers then undergo metric computation and refusal detection.}
    \label{fig:process}
\end{figure}

@@

@@

Since it is important to understand if models can provide an indication of whether their answer might likely be incorrect, we prompted some of our top performing \glspl{llm} to return the confidence in providing a correct answer on an ordinal scale.
This is similar to the verbalized confidence scores reported by \textcite{xiong2023llms}.
\Cref{fig:confidence_score_distributions} plots the distribution of those scores.
We find that the models show different distributions of confidence scores, which, for some, are skewed to the extremes.

% \begin{figure}[htb]
%     \centering
%     \includegraphics{figures/confidence_score_distributions.pdf}
%     \caption{\textbf{Distribution of confidence scores reported by \glspl{llm}.} \Glspl{llm} show different distributions of confidence scores. The confidence scores are reported on an ordinal scale from 1 to 5, with 1 indicating low confidence and 5 indicating high confidence. The bar plots show how many questions were answered with each confidence score.}
%     \label{fig:confidence_score_distributions}
%     \script{plot_confidence_score_distributions.py}
% \end{figure}

>>>>

\Cref{fig:cb_humanset} shows the distribution of topics and required skills in the human subset of the \chembench corpus.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/cb_humanset_v3.pdf}
    \caption{\textbf{Composition of the human subset.} The circular plot shows the distribution of topics and required skills in the human subset of the \chembench corpus. The human subset is a representative subset of the full corpus, with a balanced distribution of topics and skills.}
    \label{fig:cb_humanset}
\end{figure}

@@

\begin{xltabular}{\textwidth}{p{2.7cm}p{4em}X}
        \caption{\textbf{Examples for each of the topics considered in the evaluation of the \chembench corpus.} The table shows the percentage of questions in the corpus that belong to each topic, as well as example questions.}

@@

@@

\end{xltabular}

@@

\begin{xltabular}{\textwidth}{p{2.7cm}p{4em}X}
        \caption{\textbf{Examples for each of the required skills considered in the \chembench corpus.} The table shows the number of questions for each skill and an example question.}

        \toprule
        Required Skills & Number of Questions & Example Question \\
        \midrule
        Knowledge & 770 & Which of the following salts is an acidic salt? \\
        & & A. \ce{NH4Cl} \\
        & & B. \ce{Na2CO3} \\
        & & C. \ce{NaH2PO4} \\
        & & D. \ce{Zn(OH)Cl} \\
        \midrule
        Reasoning & 825 & How can the optical impact of quantum confinement on the electronic structure of a quantum dot be observed experimentally? \\
        & & A. By measuring atomic force microscopy \\
        & & B. By measuring fourier transform infrared spectroscopy \\
        & & C. By measuring photoluminescence spectrum \\
        & & D. By measuring the absorption spectroscopy \\
        \midrule
        Intuition & 1001 & Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
        & & A. [START\_SMILES]CC1(C)Oc2ccc([N+](=O)[O-])cc2[C@@H](N2CCOCC2)[C\@\@H]1O[END\_SMILES] \\
        & & B. [START\_SMILES]Cc1ccccc1N=C(S)N1CCC(NC(=O)c2ccco2)CC1[END\_SMILES] \\
        \midrule
        Calculation & 91 & Given that the average molar mass of the polymer chains in this sample of poly(lactic acid) (PLA) is \SI{595}{g mol^{-1}} using end-group analysis, where \SI{0.1619}{g} of PLA was dissolved in \SI{25}{cm^3} of benzyl alcohol and titrated with \SI{0.0400}{mol dm^{-3}} \ce{KOH} solution. The volume of \ce{KOH} solution required to reach the endpoint was \SI{6.81}{cm^3}. What is the average number of monomer units in each polymer chain of this sample? \\
        \bottomrule

    \label{tab:chembench_corpus_cognitive}
\end{xltabular}

@@

\clearpage
\subsection{Model performance}
We also evaluated the model performance on the entire \chembench corpus.
\Cref{fig:barplot_all_correct_all_questions} shows the fraction of questions that were answered correctly by the models.
Note that this ranking differs from the one on the \enquote{tiny} subset

@@

\clearpage
\subsection{Implementation}
An overview of the benchmarking pipeline implemented in \chembench is shown in \Cref{fig:process}. More detailed information can be found in the online documentation of the \chembench package at \url{https://lamalab-org.github.io/chem-bench/}.
\begin{figure}
    \centering
    \includegraphics[height=.8\textheight]{figures/process.pdf}
    \caption{\textbf{Overview of the benchmarking pipeline implemented in \chembench.} The process begins with JSON files containing task data, which are used to create \texttt{Task} objects and initialize the \texttt{ChemBenchmark}. \texttt{Prompter} objects are then created to handle different types of prompts for instruction-tuned and completion models.
    The \texttt{Task} Iterator differentiates between \gls{mcq} and other question types. For each task type, appropriate prompts are generated and passed to the \gls{llm}. The responses are then processed using regex-based parsing methods specific to \gls{mcq} or numeric answers (after obtaining the relevant part of the response from the instruction-tuned models).
    The regex-based parsing is elaborate and also able to handle special cases such as scientific notation, or roman numerals.
    If the initial parsing is unsuccessful, the system employs an \gls{llm}-based extraction method as a fallback. The parsed or extracted answers then undergo metric computation and refusal detection.}
    \label{fig:process}
\end{figure}

@@

\paragraph{Tool use}
In our study, humans were allowed to use tools for answering some questions.
They could also report what tools they used for answering questions. As \Cref{fig:tool_use} shows, the most commonly tool was some form of web search (which, according to the freetext responses, often was a multi-step process).

@@

Since it is important to understand if models can provide an indication of whether their answer might likely be incorrect, we prompted some of our top performing \glspl{llm} to return the confidence in providing a correct answer on an ordinal scale.
This is similar to the verbalized confidence scores reported by \textcite{xiong2023llms}.
We find that the models show different distributions of confidence scores, which, for some, are skewed to the extremes.

In addition, we also analyzed the confidence estimated via the log probabilities of the answer tokens. This probability of a token given the context is not necessarily the same as the confidence in the correctness of the answer. However, it is still often used as a proxy.

Our analysis of both log probabilities and prompting confidence reveals distinct calibration behaviors across different language models.
\GPTFourO demonstrates an overconfident tendency, often assigning high probabilities even to incorrect answers.
However, when \GPTFourO displays high confidence, it accurately predicts correct answers approximately 80\% of the time. In contrast, \LlamaThreeOneEightBInstruct confidence distribution is more evenly spread, with a majority of predictions centered around 0.5. High-confidence predictions from \LlamaThreeOneEightBInstruct are less frequent compared to \GPTFourO, and unlike \GPTFourO, high confidence does not necessarily correlate with a higher chance of correct answers.
%\begin{figure}[htb]
%    \centering
%    \includegraphics{figures/calibration_plot_overall_filtered.pdf}
%    \caption{\textbf{Calibration between linear probability of \glspl{llm} and correctness.} For this analysis we obtained the linear probability from the logprobs of the models. Only logprobs of the tokens corresponding to the answers were considered. Linear proabability was computed by taking exponential of logprobs (for sequences with multiple tokens, values were multiplied).  The plot shows the average predicted probability and the fraction of correct answers for each bin of linear probabilities. The ideal scenario is a diagonal line, indicating perfect calibration where the model's confidence aligns perfectly with the actual correctness. The ECE (Expected Calibration Error) value quantifies the overall calibration performance, with a lower ECE indicating better calibration.}
%    \label{fig:confidence_score_distributions}
%    \script{plot_logprobs.py}
%\end{figure}
